Mapping mutational effects along the evolutionary landscape of HIV envelope

### 2017-12-12 paper/manuscript.tex
Add file with BG505 codon mutagenesis primers
added Sarah Hilton comments

Here we use an improved version of a previously described deep mutational scanning strategy~\citep{haddox2016experimental} to measure the effects on viral replication of all single amino-acid mutations to two transmitted-founder virus Envs that differ by $>$100 mutations.
We compare these complete maps of mutational effects to identify sites that have shifted in their amino-acid preferences between the Envs.
Most sites show no detectable shifts, but 30 sites have clearly altered their preferences.
These shifted sites usually prefer a specific amino acid in one Env but have shifted to tolerate many amino acids in the other Env.
The shifted sites cluster in structure, but are often distant from any amino-acid substitutions that distinguish the two Envs, demonstrating the action of long-range epistasis.
By aggregating our measurements for both Envs, we identify sites that evolve faster or slower in nature than expected given the functional constraints measured in the lab, probably due to pressure for immune evasion.
Overall, our work provide complete across-strain maps of mutational effects that inform the analysis of Env's evolution.

@@

Both BG505 and BF520 are from clade A of the major (M) group of HIV-1.
Figure~\ref{fig:tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
BG505 and BF520 are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across clade A (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.

@@

The deep mutational scanning effectively selected for functional Env as evidenced by strong purifying selection against stop codons.
Figure~\ref{fig:mutfreqs}A shows the average frequency of mutations across Env in the plasmid mutant libraries, the mutant viruses, and wildtype controls as determined from the deep sequencing.
The mutant viruses show clear selection against stop codons and many nonsynonymous mutations (Figure~\ref{fig:mutfreqs}A).
This selection is more apparent if we correct for the background error rates estimated from the wildtype controls (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}).
The error-corrected frequencies of stop codons drop to 3\%--16\% of their original values (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}), with the residual stop codons probably due to some non-functional virions surviving due to complementation by other co-infecting virions. 
The error-corrected frequencies of nonsynonymous mutations also drop substantially (43\%--49\% of their original values), whereas the frequencies of synonymous mutations drop only slightly (85\%--95\% of their original values).
These trends are consistent with the fact that nonsynonymous mutations are often deleterious, whereas synonymous mutations often~\citep[although certainly not always, see][]{zanini2013quantifying} have only mild effects on viral replication.
Of course, Figure~\ref{fig:mutfreqs}A only summarizes one aspect of the deep mutational scanning data.
Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.

@@

A more qualitative way to assess if the deep mutational scanning authentically describes selection on Env function is to visually compare the measurements with existing knowledge.
Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} show the across-replicate average of the amino-acid preferences for each Env after re-scaling by the stringency parameters in Table~\ref{tab:phydms}.
At sites of known functional importance, these preferences are usually consistent with prior knowledge.
For instance, residues T257, D368, E370, W427, and D457 are important for Env binding to CD4~\citep{olshevsky1990identification}, and all these amino acids are highly preferred in our deep mutational scanning (Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}).
Likewise, Env has 10 disulfide bonds (linking sites 54-74, 119-205, 126-196, 131-157, 218-247, 228-239, 296-331, 378-445, 385-418, and 598-604), most of which are important for function~\citep{van2008only} -- and the cysteines at these sites are highly preferred in our deep mutational scanning.
The deep mutational scanning is also consistent with prior knowledge about sites that are tolerant of mutations.
For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure} -- and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}, such as 132 to 195).
It is beyond the scope of this paper to catalog associations between our measurements and all other prior mutational studies of Env, but we anticipate that the complete maps of mutational effects in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} will be useful for future sequence-structure-function studies.

@@

What distinguishes the sites that have undergone significant shifts?
Figure~\ref{fig:shiftfeatures}A shows the locations of the shifted sites on the crystal structure of Env.
There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core, and a statistical analysis (Figure~\ref{fig:shiftfeatures}B) finds no association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted.
However, visual inspection of Figure~\ref{fig:shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is the case (Figure~\ref{fig:shiftfeatures}C).
Therefore, the factors that drive shifts in Env's mutational tolerance often affect physically interacting clusters of residues in a coordinated fashion.

An obvious hypothesis is that strongly shifted sites have substituted between BG505 and BF520, or physically contact such substitutions.
According to this hypothesis, substitutions would alter the local physicochemical environment of the substituted site and its neighbors, thereby shifting the amino-acid preferences of sites in the physical cluster.
But surprisingly, the typical magnitude of shifts is not significantly larger at sites that have substituted or contact substitutions compared to other sites in Env (Figure~\ref{fig:shiftfeatures}C).
There is a borderline trend for the significantly shifted sites to be more likely to have substituted between BG505 and BF520 (Figure~\ref{fig:shiftfeatures}-Figure~supplement~\ref{figsupp:subs_shifts}), but most shifted sites have not substituted (only 8 of the 30 shifted sites differ in amino-acid identity between the two Envs).
The lack of strong enrichment in shifts at substituted sites contrasts with previous protein-wide experimental~\citep{doud2015site} and simulation-based~\citep{pollock2012amino,shah2015contingency} studies of shifting amino-acid preferences, which found that shifts were dramatically more pronounced at sites of substitutions.
The difference may arise because these earlier studies examined proteins that are fairly conformationally static (absolutely so in the case of the simulations).
On the other hand, Env is an extremely complex and conformationally dynamic protein~\citep{munro2014conformational,ozorowski2017open}, which may increase the opportunities for long-range epistasis to enable substitutions at one site to shift the amino-acid preferences of distant sites.

@@

We determined whether each site in Env evolves faster or slower in nature than expected given three models: that evolution is purely neutral (all nonsynonymous and synonymous mutations have equivalent effects), that sites are under the protein-level constraint measured in our experiments with BG505, or that sites are under the constraint measured with BF520. 
The first model used a standard dN/dS test~\citep[the ``FEL'' method of][]{kosakovsky2005not}, whereas the other two models performed a conceptually similar test but used ExpCM that incorporate experimental measurements as described by \citet{bloom2017identification}.
The standard dN/dS model finds hundreds of sites evolve slower than expected under neutral evolution, and only a handful of sites evolve faster than expected under neutral evolution (Table~\ref{tab:phydms}).
This finding is unsurprising, since it is well known that Env is under functional constraint.
In contrast, ExpCM that test the rates of evolution relative to the experimentally measured constraints find far fewer sites that evolve slower than expected, but many more sites that evolve faster (Table~\ref{tab:phydms}).

@@

For fitting of the phylogenetic substitution models, we used \texttt{phydms}~\citep[][\url{http://jbloomlab.github.io/phydms/}]{hilton2017phydms} to optimize the substitution model parameters and branch lengths on the fixed tree topology in Figure~\ref{fig:tree}.
The Goldman-Yang (or YNGKP) model used in Table~\ref{tab:phydms} is the M5 variant described by \citet{yang2000codon}, with the equilibrium codon frequencies determined empirically using the CF3x4 method~\citep{pond2010correcting}.
For the ExpCM shown in Table~\ref{tab:phydms}, we extended the models with empirical nucleotide frequencies described in \citet{hilton2017phydms} to also allow $\omega$ to be drawn from discrete gamma-distributed categories exactly as for the M5 model.
These ExpCM with gamma-distributed $\omega$ were implemented in \texttt{phydms} using the equations provided by \citet{yang1994maximum} (see also \url{http://jbloomlab.github.io/phydms/implementation.html#models-with-a-gamma-distributed-model-parameter}).
For both the M5 model and the ExpCM with a gamma-distributed $\omega$, the number of categories in the discretized gamma distribution is four.

@@

\begin{suppfile}
\caption{
\label{suppfile:BG505primers}
This file gives the primers used for the BG505 codon mutagenesis.
\jdbcomment{Hugh, can you add this file?}
}
\end{suppfile}

>>>>

Here we use an improved version of a previously described deep mutational scanning strategy~\citep{haddox2016experimental} to measure the effects on viral replication of all single amino-acid mutations to two transmitted-founder virus Envs that differ by $>$100 mutations.
We compare these complete maps of mutational effects to identify sites that have shifted in their amino-acid preferences between the Envs.
Most sites show no detectable shifts, but 30 sites have clearly shifted preferences.
These shifted sites usually prefer a specific amino acid in one Env but have shifted to tolerate many amino acids in the other Env.
The shifted sites cluster in structure, but are often distant from any amino-acid substitutions that distinguish the two Envs, demonstrating the action of long-range epistasis.
By aggregating our measurements for both Envs, we identify sites that evolve faster or slower in nature than expected given the functional constraints measured in the lab, probably due to pressure for immune evasion.
Overall, our work provide complete across-strain maps of mutational effects that inform the analysis of Env's evolution.

@@

Both BG505 and BF520 are from clade A of the major (M) group of HIV-1.
Figure~\ref{fig:tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
BG505 and BF520 are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity).
In our experiments, we analyze only the ectodomain and transmembrane domain of Env (we exclude the signal peptide and cytoplasmic tail).
In these regions of Env, BG505 and BF520 are identical at 549 of the 616 sites (89.1\% identity) that are alignable across clade A Envs (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.

@@

The deep mutational scanning effectively selected for functional Env as evidenced by strong purifying selection against stop codons.
Figure~\ref{fig:mutfreqs}A shows the average frequency of mutations across Env in the plasmid mutant libraries, the mutant viruses, and wildtype controls as determined from the deep sequencing.
The mutant viruses show clear selection against stop codons and many nonsynonymous mutations (Figure~\ref{fig:mutfreqs}A).
This selection is more apparent if we correct for the background error rates estimated from the wildtype controls (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}).
The error-corrected frequencies of stop codons drop to 3\%--16\% of their original values (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}), with the residual stop codons probably due to some non-functional virions surviving due to complementation by other co-infecting virions. 
The error-corrected frequencies of nonsynonymous mutations also drop substantially (43\%--49\% of their original values), whereas the frequencies of synonymous mutations drop only slightly (85\%--95\% of their original values).
These trends are consistent with the fact that nonsynonymous mutations are often deleterious, whereas synonymous mutations often~\citep[although certainly not always, see][]{zanini2013quantifying} have only mild effects on viral replication.
Figure~\ref{fig:mutfreqs}A only summarizes one aspect of the deep mutational scanning data, but Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.

@@

A more qualitative way to assess if the deep mutational scanning authentically describes selection on Env function is to visually compare the measurements with existing knowledge.
Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} show the across-replicate average of the amino-acid preferences for each Env after re-scaling by the stringency parameters in Table~\ref{tab:phydms}.
Note that throughout the rest of this paper, we use preferences re-scaled by these stringency parameters, with the re-scaling done as described in \citet{hilton2017phydms}.
At sites of known functional importance, these preferences are usually consistent with prior knowledge.
For instance, residues T257, D368, E370, W427, and D457 are important for Env binding to CD4~\citep{olshevsky1990identification}, and all these amino acids are highly preferred in our deep mutational scanning (Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}).
Likewise, Env has 10 disulfide bonds (linking sites 54-74, 119-205, 126-196, 131-157, 218-247, 228-239, 296-331, 378-445, 385-418, and 598-604), most of which are important for function~\citep{van2008only} -- and the cysteines at these sites are highly preferred in our deep mutational scanning.
The deep mutational scanning is also consistent with prior knowledge about sites that are tolerant of mutations.
For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure} -- and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}, such as 132 to 195).
It is beyond the scope of this paper to catalog associations between our measurements and all other prior mutational studies of Env, but we anticipate that the complete maps of mutational effects in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} will be useful for future sequence-structure-function studies.

@@

What distinguishes the sites that have undergone significant shifts?
Figure~\ref{fig:shiftfeatures}A shows the locations of the shifted sites on the crystal structure of Env.
There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core, and a statistical analysis (Figure~\ref{fig:shiftfeatures}B) finds no association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted.
However, Figure~\ref{fig:shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is the case (Figure~\ref{fig:shiftfeatures}C).
Therefore, the factors that drive shifts in Env's mutational tolerance often affect physically interacting clusters of residues in a coordinated fashion.

An obvious hypothesis is that strongly shifted sites have substituted between BG505 and BF520, or physically contact such substitutions.
According to this hypothesis, substitutions would alter the local physicochemical environment of the substituted site and its neighbors, thereby shifting the amino-acid preferences of sites in the physical cluster.
But surprisingly, the typical magnitude of shifts is not significantly larger at sites that have substituted, or at sites that contact sites that have experienced substitutions (Figure~\ref{fig:shiftfeatures}C).
There is a borderline trend for the significantly shifted sites to be more likely to have substituted between BG505 and BF520 (Figure~\ref{fig:shiftfeatures}-Figure~supplement~\ref{figsupp:subs_shifts}), but most shifted sites have not substituted (only 8 of the 30 shifted sites differ in amino-acid identity between the two Envs).
The lack of strong enrichment in shifts at substituted sites contrasts with previous protein-wide experimental~\citep{doud2015site} and simulation-based~\citep{pollock2012amino,shah2015contingency} studies of shifting amino-acid preferences, which found that shifts were dramatically more pronounced at sites of substitutions.
The difference may arise because these earlier studies examined proteins that are fairly conformationally static (absolutely so in the case of the simulations).
On the other hand, Env is an extremely complex and conformationally dynamic protein~\citep{munro2014conformational,ozorowski2017open}, which may increase the opportunities for long-range epistasis to enable substitutions at one site to shift the amino-acid preferences of distant sites.

@@

We determined whether each site in Env evolves faster or slower in nature than expected given three models: that evolution is purely neutral (all nonsynonymous and synonymous mutations have equivalent effects), that sites are under the protein-level constraint measured in our experiments with BG505, or that sites are under the constraint measured with BF520. 
The first model used a standard dN/dS test~\citep[the ``FEL'' method of][]{kosakovsky2005not}, whereas the other two models performed a conceptually similar test but used ExpCM that incorporate experimental measurements as described by \citet{bloom2017identification}.
The standard dN/dS model finds hundreds of sites that evolve slower than expected under neutral evolution (Table~\ref{tab:phydms}, $\omega_r < 1$), and only a handful of sites that evolve faster than expected under neutral evolution (Table~\ref{tab:phydms}, $\omega_r > 1$).
This finding is unsurprising, since it is well known that Env is under functional constraint.
In contrast, ExpCM that test the rates of evolution relative to the experimentally measured constraints find far fewer sites that evolve slower than expected, but many more sites that evolve faster (Table~\ref{tab:phydms}).

@@

For fitting of the phylogenetic substitution models, we used \texttt{phydms}~\citep[][\url{http://jbloomlab.github.io/phydms/}]{hilton2017phydms} to optimize the substitution model parameters and branch lengths on the fixed tree topology in Figure~\ref{fig:tree}.
The Goldman-Yang (or YNGKP) model used in Table~\ref{tab:phydms} is the M5 variant described by \citet{yang2000codon}, with the equilibrium codon frequencies determined empirically using the CF3x4 method~\citep{pond2010correcting}.
For the ExpCM shown in Table~\ref{tab:phydms}, we extended the models with empirical nucleotide frequencies described in \citet{hilton2017phydms} to also allow $\omega$ to be drawn from discrete gamma-distributed categories exactly as for the M5 model.
These ExpCM with gamma-distributed $\omega$ were implemented in \texttt{phydms} using the equations provided by \citet{yang1994maximum} (see also \url{http://jbloomlab.github.io/phydms/implementation.html#models-with-a-gamma-distributed-model-parameter}).
The preferences were re-scaled by the stringency parameters in Table~\ref{tab:phydms} as described in \citet{hilton2017phydms}.
For both the M5 model and the ExpCM with a gamma-distributed $\omega$, we used four categories for the discretized gamma distribution.

@@

\begin{suppfile}
\caption{
\label{suppfile:BG505primers}
The sequences of the primers used for the BG505 codon mutagenesis are in the file \texttt{BG505\_codon\_mutagenesis\_primers.txt}.
}
\end{suppfile}
